---
title: "Dissecting tumor transcriptional heterogeneity from multi-tumor single-cell RNA-seq data using GBCD"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    toc_float: no
    highlight: textmate
    theme: readable
vignette: >
  %\VignetteIndexEntry{Dissecting tumor transcriptional heterogeneity from multi-tumor single-cell RNA-seq data using GBCD}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Overview
--------

This vignette shows how to apply GBCD ("Generalized Binary Covariance
Decomposition") to jointly analyze single-cell RNA-seq (scRNA-seq)
data from *malignant cells* collected from multiple patients and/or
studies.

GBCD can effectively dissect tumor transcriptional heterogeneity into
patient/study-specific and shared gene expression programs
(GEPs). GBCD is "unsupervised" in that, unlike tumor-by-tumor (e.g.,
[Puram et al. 2017][hnscc]) and many harmonization approaches (e.g.,
[Harmony][harmony], [LIGER][liger]), it does not use information about
which cell comes from which tumor or study. Instead, GBCD only
requires the combined scRNA-seq data from all tumors, which are stored
as an $N \times J$ matrix $Y$ of expression values with entries
$y_{ij}$, where $i=1,\dots,N$ indexes malignant cells and
$j=1,\dots,J$ indexes genes. In typical applications, $Y$ contains 
"shifted log" counts (following the terminology of
[this paper][ahlmann-eltze-huber-2023]).

GBCD yields a decomposition of the expression data matrix $Y$ into
matrices $L$ and $F$ such that $Y \approx L F^T$:
$$
y_{ij} \approx \sum_{k=1}^K l_{ik} f_{jk}.
$$
The elements $l_{ik} \geq 0$ each represent the membership of cell $i$
in GEP $k$; $f_{jk}$ represents the effect of GEP $k$ on the
expression of gene $j$. When $y_{ij}$ are shifted log counts, each
$f_{jk}$ approximately represents the log-fold change (LFC) associated
with membership in GEP $k$, so we refer to the $f_{jk}$ values as
LFCs, and we refer to the vector of LFCs $(f_{1k}, \dots, f_{Jk})^T$
as the "signature" of GEP $k$.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#", collapse = TRUE, results = "hold",
                      fig.align = "center", dpi = 120)
```

We begin our analysis by loading the needed R packages.

```{r load-pkgs, message=FALSE, warning=FALSE}
library(Matrix)
library(ggplot2)
library(cowplot)
library(ggrepel)
library(pheatmap)
library(Rtsne)
library(flashier)
library(gbcd)
```

The HNSCC dataset
-----------------

This dataset contained gene expression data for 2,176 malignant
cells collected from primary tumors from 10 HNSCC patients and
matching lymph node (LN) metastases from 5 of these
patients. [Puram et al. (2017)][hnscc] found that each of these 10
patients clearly mapped to a molecular subtype of HNSCC, whose
signatures were previously defined by analysis of bulk expression data
of 279 [TCGA][tcga] HNSCC tumors. [Puram et al. (2017)][hnscc] defined the
transformed counts as $y_{ij} = \log_2(1 + \mathrm{TPM}_{ij}/10)$,
where $\mathrm{TPM}_{ij}$ was the transcript-per-million (TPM) value
for gene $j$ in cell $i$.

```{r load-data}
data(hnscc)
dim(hnscc$Y)
head(hnscc$info)
```

Visualize HNSCC data using t-SNE
--------------------------------

We plot the 2-D *t*-SNE embedding of the 2,176 malignant cells,
colored by patient-of-origin and tumor stage (primary tumor, LN
metastasis), and shaped is varied by tumor molecular subtype. Due to
the presence of strong inter-tumor heterogeneity, these cells
demonstrate strong patient effects that are typical of cancer data;
the major structure in the t-SNE visualization is the clustering of
the cells by patient:

```{r plot-tSNE, fig.height=4, fig.width=5}
set.seed(1)
cols <- order(apply(hnscc$Y,2,sd),decreasing = TRUE)
cols <- cols[1:3000]
res  <- Rtsne(as.matrix(hnscc$Y[,cols]),normalize = TRUE)
colnames(res$Y) <- c("tsne1","tsne2")
pdat <- cbind(res$Y,hnscc$info)
ggplot(pdat,aes(x = tsne1,y = tsne2,color = sample,shape = subtype)) +
  geom_point(size = 1) + scale_color_manual(values = hnscc$sample_col) +
  scale_shape_manual(values = c(15,16,18)) +
  theme_cowplot(font_size = 10)
```

Estimate GEP memberships and signatures using GBCD
--------------------------------------------------

To fit a GBCD model, "Kmax" needs to set. It is an upper bound of
the number of components, $K$. This is just for initialization
purposes; the final number of GEPs returned by GBCD is close to but
often not equal to (and can be larger than) Kmax. Generally speaking,
a larger Kmax allows for identifying finer structure in tumor
transcriptional heterogeneity, at the expense of higher computational
cost. A good choice of Kmax depends on the complexity of the analyzed
data (e.g., the number of patients and studies contained), and is
often unknown in advance. Based on our experience, a value of Kmax
between 10 and 40 usually works well, and users are encouraged to
explore and compare the results for different values of Kmax.

Users also need to specify a prior for GEP membership values. GBCD
assigns a "generalized binary" (GB) prior independently to each entry
of $L$, $$l_{ik}\sim (1-\pi_k^l) \delta_0 + \pi_k^l \: N_{+} (\mu_k,
\sigma_k^2),$$ where the ratio $\sigma_k/\mu_k$ is fixed at some
pre-specified small value. In this application, we set $\sigma_k/\mu_k
= 0.04$ for each $k$. A smaller ratio is generally recommended for
larger datasets containing more cells, and a value between 0.01 to 0.1
usually works well.

Note that the patient/dataset identity information of cells is *not* 
needed to run GBCD. 

For the purposes of reducing the running time, here we set `Kmax =
12`, and we limit the number of backfitting iterations to `maxiter1 =
100` (for initializing L), `maxiter2 = 50` (for estimating L), and
`maxiter3 = 50` (for estimating F).

```{r fit-gbcd, warning=FALSE}
res.gbcd <- fit_gbcd(Y = hnscc$Y, Kmax = 12, maxiter1 = 100,
                     maxiter2 = 50, maxiter3 = 50, 
					 prior = flash_ebnm(prior_family = "generalized_binary",
					                    scale = 0.04))
```

Note that users can also specify other nonnegative priors for GEP
membership values. For example, users can specify a point exponential
prior by setting `prior = ebnm_point_exponential` in the call to
`fit_gbcd`. However, we find that the GB prior usually performs the
best empirically to dissect tumor transcriptional heterogeneity from
multiple samples and/or studies.

Note that the above call to `fit_gbcd` may take 10 minutes or slightly
longer to complete on a laptop with 24GB RAM. 


Interpret GEP membership estimates
------------------------

We now look at the GEP membership estimates $L$, in which elements
$l_{ik}$ represent membership of cell $i$ in GEP $k$. For example
these are the memberships for the first few cells:

```{r summary-L}
head(round(res.gbcd$L, 3))
```

Let's create a heatmap to visualize GEP membership estimates. First we
add the sample and molecular subtype annotation of cells to visualize
the GEP memberships:

```{r plot-L-1}
anno <- data.frame(sample = hnscc$info$sample, subtype = hnscc$info$subtype)
rownames(anno) <- rownames(res.gbcd$L)
anno_colors <- list(sample = hnscc$sample_col, subtype = hnscc$subtype_col)
cols <- colorRampPalette(c("gray96", "red"))(50)
brks <- seq(0, 1, 0.02)
```

Now we can plot the annotated heatmap of GEP memberships:

```{r plot-L-2, fig.height=5, fig.width=6}
pheatmap(res.gbcd$L[order(anno$sample), -c(1)], cluster_rows = FALSE,
         cluster_cols = FALSE, show_rownames = FALSE, annotation_row = anno,
		 annotation_colors = anno_colors, annotation_names_row = FALSE,
		 angle_col = 45, fontsize = 9, color = cols, breaks = brks,
		 main = "")
```

This heatmap shows the memberships of the 2,176 cells (rows) for the
20 GEPs (the "baseline" is not included). Cells are arranged
top-to-bottom by tumor molecular subtype and by patient-of-origin.
Membership values are rescaled separately for GEPs so that the maximum
membership value for each GEP is always 1.

From this heatmap, we see that GEPs 1-3 correspond closely to the
molecular subtypes previously defined based on bulk RNA-seq data: GEP1
is largely active only in cells from the 2 classical patients; GEP2 is
mainly active in cells from the 7 basal patients; and GEP3 is mainly
active in cells from the one atypical patient. This demonstrates that
GBCD can extract the molecular subtype information *de novo* from the
single cell data alone.

Among other GEPs, some are active predominantly in an individual
patient and can thus be interpreted as patient-specific GEPs, such as
GEPs 8, 9 and 11; other GEPs are shared across patients but do not
correspond to molecular subtypes. The degree of patient-specific
vs. shared expression for each GEP can be quantified using the
"gep_annotation" function.

GEP signature estimates
-----------------------

We next look at the matrix of GEP signature estimates $F$ with each
entry $f_{jk}$ approximately representing the log-fold
change associated with membership in GEP $k$ (with the base-2 logarithm).

```{r summary-lfc}
head(round(res.gbcd$F$lfc, 3))
```

Here we create a volcano plot to visualize the gene signature for a
given GEP (using GEP2 as an example). This plot makes use of
posterior statistics outputted by the `fit_gbcd` call, including the z-scores
and the local false sign rates ([*lfsr*][lfsr]).

```{r plot-lfc, fig.height=5, fig.width=6}
pdat <- data.frame(gene = rownames(res.gbcd$F$lfc), 
                   lfc = res.gbcd$F$lfc[, "GEP2"], 
                   z = abs(res.gbcd$F$z_score[, "GEP2"]), 
                   lfsr = res.gbcd$F$lfsr[, "GEP2"],
                   stringsAsFactors = FALSE)
pdat <- transform(pdat, lfsr = cut(lfsr, c(-1, 0.001, 0.01, 0.05, Inf)))
rows  <- with(pdat, which(!(abs(lfc) > quantile(abs(lfc), 0.998) | (z > 10))))
pdat[rows, "gene"] <- ""
ggplot(pdat, aes(x = lfc, y = z, color = lfsr, label = gene)) + geom_point() + 
  geom_text_repel(color = "black", size = 2.5, segment.color = "black",
                  segment.size = 0.25, min.segment.length = 0,
				  max.overlaps = Inf, na.rm = TRUE) +
  scale_color_manual(values = c("coral", "orange", "gold", "deepskyblue")) +
  labs(x = "log-fold change", y = "|posterior z-score|") + 
  guides(colour = guide_legend(override.aes = list(size = 2))) + 
  theme(plot.title      = element_text(hjust = 0.5,size = 12),
        axis.text       = element_text(size = 10),
		axis.title      = element_text(size = 10),
		legend.title    = element_text(size = 12),
		legend.text     = element_text(size = 10),
		legend.position = "bottom") +
  ggtitle("Volcano plot of gene signature for GEP2")
```

Session info
------------

This was the version of R and the packages that were used to generate
these results.

```{r session-info}
sessionInfo()
```

[ahlmann-eltze-huber-2023]: https://doi.org/10.1038/s41592-023-01814-1
[hnscc]: https://doi.org/10.1016/j.cell.2017.10.044
[harmony]: https://portals.broadinstitute.org/harmony/
[liger]: https://github.com/welch-lab/liger/
[tcga]: https://doi.org/10.1038/nature14129
[ebnm]: https://stephenslab.github.io/ebnm/
[gbcd]: https://github.com/stephenslab/gbcd/
[lfsr]: https://doi.org/10.1093/biostatistics/kxw041
